\documentclass{article}
\usepackage[margin=2cm,bottom=2cm]{geometry}
\usepackage{hyperref}
\usepackage{comment}
\usepackage[utf8]{inputenc}
\usepackage{graphicx}
\usepackage{mathtools}
\usepackage[normalem]{ulem}
\usepackage{color}
\usepackage{setspace}
\usepackage{MnSymbol}
\usepackage[T1]{fontenc}
\usepackage{soul}

\newcommand\tab[1][1cm]{\hspace*{#1}}
\DeclareMathSizes{10}{10}{10}{10}

\begin{document}
\title{COMP/CMPE 314 - Principles of Programming Languages - Notes}
\author{Chris Stephenson, Istanbul Bilgi University, Department of Mathematics, and course students}
\maketitle

\section*{The "real" world}
\begin{flushleft}
\textbf{"Evaluation \underline{Strategies}"}\\
\begin{itemize}
 \item a strange word to use?
 \item we make the languages.
\end{itemize}
The problem is that most languages have design defects, so evaluating them is confusing.\\
\bigskip
\textbf{What do we do?}\\
When and What and How parameters are evaluated?
\end{flushleft}

\subsection{When?}
In Java;
\begin{verbatim}
  myfunc(a, g(b), c + d);
  myFunc(a++, a++, a++);	// order matters!
\end{verbatim}
** Java compiler (javac) starts evaluating from \underline{left} however C compiler (gcc) starts from \underline{right}.**\\
\bigskip

\textbf{Are the parameters evaluated left to right or right to left?}\\
In Java, left to right\\
In C, \underline{UNDEFINED!}? (Different compilers different answers)


\begin{flushleft}
\textbf{\underline{OR}} We cannot evaluate the parameters at all. And simply pass the expressions to the function body. (All programming languages do this for \underline{some} things)\\
Racket is a \underline{strict} language, which means parameters \underline{are} evaluated.
\end{flushleft}

\begin{verbatim}
;#lang lazy
#lang racket
(define (my-if condition t f)
  (cond
    (condition t)
    (else f)))
    
(define (fac n)
  (my-if (< n 1) 1 (* n (fac (- n 1)))))
\end{verbatim}
Here \verb|my-if| looks like a function. But it isn't a function in racket. It tries to calculate parameters first. Thus it leads to infinitive loop.\\
If we uncomment \verb|#lang lazy| it will work. 

\begin{flushleft}
\underline{Except} for \underline{if}\\
In Java \verb|f(a) && g(b)| is a boolean \st{expression}\\
They call it "Short circuited evaluation"\\
Evaluated left to right and stop when we know the answer.
\end{flushleft}

\begin{flushleft}
 
\underline{A practical example:}
\begin{verbatim}
  int i = 0;
  while(i < a.length && a[i] != 0) {
    ...
    i++;
  }
  
  while(a[i]!=0 && i<a.length){	// after a swap, what goes wrong?
    ...
    i++;
  }
\end{verbatim}
If the condition goes to \verb|i > a.length|, the \verb|a[i]| throws an error. So the order is important!
\end{flushleft}

\begin{flushleft}
Is evaluation in this language "Strict?" "Non Strict?" "Sometimes Strict?"\\
A different way of binding formal parameters: \noindent\rule{3cm}{0.4pt}
\end{flushleft}
\begin{center}
\textbf{\underline{Environments}}
\end{center}

\begin{flushleft}
\begin{itemize}
 \item Substitution is not how the real world works (For reasons of efficiency)
 \item An environment is an ordered store of identifier/value pairs
 \item For function application we add formal parameter/actual parameter value pairs to the environment and use that environment to evaluate the function body.
 \item Our \underline{evaluator} takes the environment as an additional parameter when the evaluator evaluates an identifier, it looks up the value in the environment.
\end{itemize}

\end{flushleft}
\subsection*{Data Definitions}
An environment is empty \underline{OR} an identifier/value pair plus an environment.
\begin{itemize}
\item look up Env. environment identifier $\rightarrow$ value or \underline{Error}
\item extend Env. environment identifier $\rightarrow$ environment
\end{itemize}
\begin{verbatim}
Example: 
(fundef myFunc (a b c) (+ a (* b c))) //function definition

(myFunc 3 7 (+ 2 9)) //function application
\end{verbatim}

\subsubsection*{Evaluate a function application}
\begin{verbatim}
(eval (extend Env (extend Env (extend Env (empty Env, 'a, 3) 'b, 7) 'c, 11) '(+ a (* b c))))

(fundef f1 x (f2 4))
(fundef f2 y (+ x y))
\end{verbatim}
Whats wrong with f2?\\
It's a free identifier!\\
In Java:
\begin{verbatim}
int f2(int y){
  return x + y;
}
\end{verbatim}
evaluate (f1 42)\\
 x gets bound to 42\\
 y gets bound to 4\\
 (+ x y) looks up these values in the environment\\
 The answer is 46 \textbf{Wrong!!}\\
\bigskip
Identifier capture!\\
This is called "dynamic scope" and it is a bug.
\bigskip
\begin{flushleft}
\underline{The solution} (for now)\\
When we apply a function, do \underline{not} extend the \underline{current environment} with formal parameter/value bindings. Extend the \underline{empty} environment.
\end{flushleft}

\subsubsection*{Grammar}
\begin{flushleft}
$1)$ $\Lambda\rightarrow\upsilon$

$2)$ $\Lambda\rightarrow(\lambda$ 
$\upsilon$
$\Lambda)$

$3)$ $\Lambda\rightarrow(\Lambda$
$\Lambda)$

Data structure and therefore definitions and programs will reflect this.\\
\bigskip
M[x:=N]\\
\end{flushleft}

\begin{flushleft}
$1)$ M is an identifier\\
\tab{a) M=x $\Rightarrow$ M[x	:=N] is N}\\
\tab{b) M$\neq$x $\Rightarrow$ M[x:=N] is M}\\
\bigskip
$2)$ M is an application (just do the two parts)\\
\bigskip
$3)$ M is a function definition\\
\tab{Case (a) is easy $\lambda$yM1, y=x}\\
\tab{Case (b) is hard $\lambda$yM1, y$\neq$x}\\
\end{flushleft}

\subsubsection*{Church Numerals}
\begin{flushleft}
n $\equiv$ ($\lambda$ f ($\lambda$ x (f (f (f...x)))))\\
$\tab{\tab{\tab{\overline{n f's}}}}$
\end{flushleft}

\begin{flushleft}
ChurchB(n) $\equiv$ (f (ChurchB(n -1)))\\
ChurchB(1) $\equiv$ ($\lambda$ f ($\lambda$ x (f x)))\\
\bigskip
Church(n) = ($\lambda$ f (x x (ChurchB(n))))\\
\end{flushleft}

\begin{flushleft}
TWO=($\lambda$ f ($\lambda$ x (f (f x))))\\
ZERO=($\lambda$ f ($\lambda$ x x))\\
\end{flushleft}
\bigskip
\begin{flushleft}
\textbf{A successor function "SUCC"}\\
SUCC=($\lambda$ n ($\lambda$ f ($\lambda$ x (f ((n f) x))))\\
onSUCC=($\lambda$ n ($\lambda$ f ($\lambda$ x ((n f) (f x))))
\end{flushleft}

\begin{flushleft}
To add m and n;\\
($\lambda$ m ($\lambda$ n ((m succ) n)))
\end{flushleft}

\end{document}
  